---
title: "vis"
author: "kate whelan - 22386463"
date: "2026-02-24"
output:
  pdf_document: default
  html_document: default
---
```{r}
library(ggplot2)
library(tidyverse)
```



```{r}
library(tidyverse)
library(lubridate)

nics_firearm <- 
  nics_firearm_background_checks %>%
  mutate(month = ymd(paste0(month, "-01")))

nics_firearm <- nics_firearm |> drop_na()
view(nics_firearm)

```

```{r}
permit_avg <- nics_firearm |>
  group_by(state) |>
  summarise(avg_monthly_permits = mean(permit))
  

ggplot(permit_avg,
       aes(x =  reorder(state, avg_monthly_permits),
           y = avg_monthly_permits)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "Average Monthly Permit Background Checks by State",
    x = "State",
    y = "Average Monthly Permit Checks"
  ) 
```

```{r}
permit_totals <- nics_checks_last_five_years |>
  group_by(state) |>
  summarise(total_permits = sum(permit))

ggplot(permit_totals, 
       aes(x = reorder(state, total_permits), 
           y = total_permits)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "Total Firearm Permit Background Checks by State",
    x = "State",
    y = "Total Permit Checks"
  )
```

```{r}
state_totals <- nics_firearm |>
  group_by(state) |>
  summarise(total_checks = sum(totals)) |>
  arrange(desc(total_checks))

ggplot(state_totals,
       aes(x = reorder(state, total_checks),
           y = total_checks)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "Total Firearm Background Checks by State",
    x = "State",
    y = "Total Checks"
  )
```

```{r}
ggplot(data = nics_firearm, aes(x = permit, y = month, colour = month)) +
         geom_point() + facet_wrap(~state)
```
```{r}
ggplot(data = nics_firearm, aes(x = permit, y = month, colour = month)) +
         geom_line() + facet_wrap(~state)
```

```{r}
library(ggplot2)
library(dplyr)
library(maps)
library(plotly)
```


#Total permits by state with scale from -1 to 1


```{r}
us_map <- map_data("state")

nics_checks_last_five_years <- nics_checks_last_five_years |>
mutate(state = tolower(state))

state_permit_totals <- nics_checks_last_five_years |>
group_by(state) |>
summarise(total_permit = sum(permit))

map_data_joined <- us_map |>
left_join(state_permit_totals, by = c("region" = "state"))

map_data <- map_data_joined |>
mutate(scaled_values = 2 * (total_permit - min(total_permit)) / 
(max(total_permit) - min(total_permit)) -1)
  

g1 <- ggplot(map_data, aes(x = long, y = lat, group = group, fill = scaled_values,
  text = paste("State:",region,"<br>Permits:", total_permit))) +
  geom_polygon(color = "white") +
  scale_fill_gradient2(
  low = "red",
  high = "blue",
  midpoint = (mean(map_data$scaled_values + 1))
  ) +
  labs(
  title = "Total Firearm Permits by State",
  fill = "total permits"
  ) +
  theme_dark() 


ggplotly(g1, tooltip = "text")


```



#Total permit rechecks by state with scale from -1 to 1

```{r}
us_map <- map_data("state")

nics_checks_last_five_years <- nics_checks_last_five_years |>
mutate(state = tolower(state))

state_permit_totals <- nics_checks_last_five_years |>
group_by(state) |>
summarise(total_permit = sum(permit_recheck))

map_data_joined <- us_map |>
left_join(state_permit_totals, by = c("region" = "state"))

map_data <- map_data_joined |>
mutate(scaled_values = 2 * (total_permit - min(total_permit)) / 
(max(total_permit) - min(total_permit)) -1)



g2 <- ggplot(map_data, aes(x = long, y = lat, group = group, fill = scaled_values,
 text = paste("State:",region,"<br>Permit rechecks:", total_permit))) +
 geom_polygon(color = "white") +
 scale_fill_gradient2(
 low = "red",
 high = "blue",
 midpoint = 0) +
 labs(
 title = "Total Firearm Background Checks by State",
 fill = "total rechecks") +
 theme_dark()


ggplotly(g2, tooltip = "text")
```



#Total permits by state with log and rank based scaling from -1 to 1
#use or percent_rank() - 
#use of log10
```{r}
us_map <- map_data("state")

nics_checks_last_five_years <- nics_checks_last_five_years |>
mutate(state = tolower(state))

state_permit_totals <- nics_checks_last_five_years |>
group_by(state) |>
summarise(total_permit = sum(permit))

map_data_joined <- us_map |>
left_join(state_permit_totals, by = c("region" = "state"))

map_data <- map_data_joined |>
mutate(scaled_values = 2 * (percent_rank(log10(total_permit + 1)) - 0.5))

g3 <- ggplot(map_data, aes(x = long, y = lat, group = group, fill = scaled_values,
 text = paste("State:",region,"<br>Permits:", total_permit))) +
 geom_polygon(color = "white") +
 scale_fill_gradient2(
 low = "red",
 mid = "white",
 high = "blue",
 midpoint = 0) +
 labs(
 title = "Total Firearm Permits by State",
 fill = "permits") +
 theme_dark() 


ggplotly(g3, tooltip = "text")

```

#Total permit rechecks by state with log and rank based scaling from -1 to 1
#use or percent_rank()
#use of log10
#think this graph is best to show the divide between states?!?!
```{r}
us_map <- map_data("state")

nics_checks_last_five_years <- nics_checks_last_five_years |>
  mutate(state = tolower(state))

state_permit_totals <- nics_checks_last_five_years |>
  group_by(state) |>
  summarise(total_permit = sum(permit_recheck))

map_data_joined <- us_map |>
  left_join(state_permit_totals, by = c("region" = "state"))

map_data <- map_data_joined |>
  mutate(
    scaled_values = 2 * (percent_rank(log10(total_permit + 1)) - .5)
  )

g4 <- ggplot(map_data, aes(x = long, y = lat, group = group, fill = scaled_values,
 text = paste("State:",region,"<br>Permit rechecks:", total_permit))) +
 geom_polygon(color = "white") +
 scale_fill_gradient2(
 low = "red",
 mid = "white",
 high = "blue",
 midpoint = 0) +
 labs(
 title = "Total Firearm Permits rechecks by State",
 fill = "permit rechecks") +
 theme_dark()

ggplotly(g4, tooltip = "text")
```

#illinois is a clear outlier from all graphs seems like an error with data input.






