---
title: "roc_change"
author: "Owen Frank O'Connor"
date: "2026-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Does the rate of change in total firearms background checks over time vary across states?

---

## 1. Libraries & Data Load

```{r load-data}
library(tidyverse)
library(lubridate)
library(scales)
library(ggrepel)
library(viridis)
library(broom)

nics <- read_csv("../data/nics-firearm-background-checks.csv") |>
  mutate(date = ym(month), year = year(date))

state_pop        <- readRDS("../data/other_sources/population/state_pop_2000_2023.RDS")
nics_participation <- readRDS("../data/other_sources/nics_participation.RDS")
```

---

## 2. Prepare Annual Per-Capita Totals

```{r annual-pop}
us_states <- c(state.name, "District of Columbia")

annual_pop <- nics |>
  filter(state %in% us_states, year >= 2000, year <= 2023) |>
  group_by(state, year) |>
  summarise(annual_total = sum(totals, na.rm = TRUE), .groups = "drop") |>
  left_join(state_pop, by = c("state", "year")) |>
  left_join(nics_participation, by = "state") |>
  mutate(checks_per_100k = (annual_total / population) * 1e5) |>
  filter(!is.na(checks_per_100k), checks_per_100k > 0)
```

## 4. Define Four ~6-Year Segments

```{r define-segments}
segment_levels <- c("2000–2005", "2006–2011", "2012–2017", "2018–2023")

annual_pop <- annual_pop |>
  mutate(
    checks_per_1k = (annual_total / population) * 1000,
    segment = case_when(
      year %in% 2000:2005 ~ "2000–2005",
      year %in% 2006:2011 ~ "2006–2011",
      year %in% 2012:2017 ~ "2012–2017",
      year %in% 2018:2023 ~ "2018–2023",
      TRUE                ~ NA_character_
    )
  ) |>
  filter(!is.na(segment), checks_per_1k > 0) |>
  mutate(segment = factor(segment, levels = segment_levels))
```

---

## 5. Fit Per-Segment Linear Trends by State

For each state–segment pair we fit:

$${\text{checks_per_1k}}_{t} \;=\; \beta_0 \;+\; {\beta_1}\cdot\text{year}_{t}$$

The slope {\beta_1} estimates the annual rate of change in background checks
per 1,000 residents within that ~6-year window.

```{r segment-slopes}
segment_slopes <- annual_pop |>
  group_by(state, segment) |>
  nest() |>
  mutate(
    fit    = map(data, \(d) lm(checks_per_1k ~ year, data = d)),
    tidied = map(fit,  \(m) tidy(m, conf.int = TRUE))
  ) |>
  unnest(tidied) |>
  filter(term == "year") |>
  select(state, segment,
         slope = estimate, se = std.error,
         conf.low, conf.high, p.value) |>
  ungroup()
```

---

## 6. Formal Hypothesis Test: Do Slopes Vary Across Segments?

**Hypotheses (per state)**

- \(H_0\): the slope of \(\text{checks_per_1k} \sim \text{year}\) is equal
  across all four segments  
- \(H_1\): at least one segment has a significantly different slope

We test this by comparing nested models via an F-test:

- *Reduced*: `checks_per_1k ~ year + segment`  (common slope, shifted intercepts)  
- *Full*: `checks_per_1k ~ year * segment`  (separate slope per segment)

```{r interaction-ftest}
ftest_results <- annual_pop |>
  group_by(state) |>
  nest() |>
  mutate(
    mod_full    = map(data, \(d) lm(checks_per_1k ~ year * segment, data = d)),
    mod_reduced = map(data, \(d) lm(checks_per_1k ~ year + segment, data = d)),
    anova_res   = map2(mod_reduced, mod_full, anova),
    anova_tidy  = map(anova_res, tidy)
  ) |>
  unnest(anova_tidy) |>
  filter(!is.na(p.value)) |>
  select(state, df, statistic, p.value) |>
  mutate(
    significant = p.value < 0.05,
    sig_label   = if_else(significant, "Significant (p < 0.05)", "Not Significant")
  ) |>
  ungroup()
```

```{r ftest-summary}
n_sig <- sum(ftest_results$significant)
n_tot <- nrow(ftest_results)

cat(sprintf(
  "%d of %d states show significantly different slopes across segments (α = 0.05).\n\n",
  n_sig, n_tot
))

ftest_results |>
  filter(significant) |>
  arrange(p.value) |>
  select(State = state, `F-statistic` = statistic, `p-value` = p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  print(n = Inf)
```

---

## 7. Visualisation

### Slope Trajectories — Significantly Varying States

Of the states where \(H_0\) is rejected, a  number of them have been idenitfied elsehwere for weird behaviour

```{r slope-trajectory, fig.width = 15, fig.height = 4}
bad_states <- c("Kentucky", "Illinois", "Utah", "Indiana", "North Carolina", "California", "North Dakota")
good_sig_states <- ftest_results |>
  filter(significant, !state %in% bad_states)

segment_slopes |>
  filter(state %in% good_sig_states$state) |>
  ggplot(aes(x = segment, y = slope, group = state, colour = state)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey55", linewidth = 0.5) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.18, alpha = 0.45, linewidth = 0.6) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 2.8) +
  facet_wrap(~ state, scales = "free_y", ncol = 5) +
  labs(x = "Segment", y = "Slope", title = "Slopes for Significant Good States") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```
